dispRity manual

### 2017-11-17 inst/gitbook/dispRity_manual.Rmd
updated manual

@@

>>>>

--- 
title: "dispRity manual"
author: "Thomas Guillerme (guillert@tcd.ie) and Natalie Cooper (natalie.cooper@nhm.ac.uk)"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
output: bookdown::gitbook
documentclass: book
link-citations: yes
github-repo: TGuillerme/dispRity
description: "dispRity R package vignette"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# `dispRity`

This is a package for measuring disparity in `R`.
It allows users to summarise matrices as representations as multidimensional spaces into a single value or distribution describing a specific aspect of this multidimensional space (the disparity).
Multidimensional spaces can be ordinated matrices from MDS, PCA, PCO, PCoA but the package is *not* restricted to any type of matrices!
This manual is based on the version `0.4.2`.

## What is `dispRity`?
This is a modular package for measuring disparity in `R`.
It allows users to summarise ordinated matrices (e.g. MDS, PCA, PCO, PCoA) to perform some multidimensional analysis.
Typically, these analysis are used in palaeobiology and evolutionary biology to study the changes in morphology through time.
However, there are many more applications in ecology, evolution and beyond.

### Modular?
Because their exist a multitude of ways to measure disparity, each adapted to every specific question, this package uses an easy to modify modular architecture.
In coding, each module is simply a function or a modification of a function that can be passed to the main functions of the package to tweak it to your proper needs!
In practice, you will notice throughout this manual that some function can take other functions as arguments: the modular architecture of this package allows you to use any function for these arguments (with some restrictions explained for each specific cases).
This will allow you to finely tune your multidimensional analysis to the needs of your specific question!

## Installing and running the package

You can install this package easily if you use the latest version of `R` (> 3.4.0) and `devtools`.

```{r, eval=FALSE}
## Checking if devtools is already installed
if(!require(devtools)) install.packages("devtools")

## Installing the latest released version directly from GitHub
install_github("TGuillerme/dispRity", ref = "release")

## Loading the package
library(dispRity)
```

```{r, eval = TRUE, echo = FALSE, message = FALSE}
## loading the package and setting up the start seed.
library(dispRity)
## Setting a random seed for repeatability
set.seed(123)
```

Note this uses the `release` branch (version`0.4`). <!-- @@@ -->
For the piping-hot (but potentially unstable) version, you can change the argument `ref = release` to `ref = master`.
`dispRity` depends mainly on the `ape` package and uses functions from several other packages (`ade4`, `geometry`, `grDevices`, `hypervolume`, `paleotree`, `snow`, `Claddis`, `geomorph` and `RCurl`).

## Why not CRAN?

This package is not available on CRAN.
This is mainly because some parts are still in development and that the reactivity of GitHub is better for implementing new suggestions from users.
However, the package follows the strict (and useful!) CRAN standards via Travis.

<!-- badges out start -->
[![Build Status](https://travis-ci.org/TGuillerme/dispRity.svg?branch=release)](https://travis-ci.org/TGuillerme/dispRity)
[![codecov](https://codecov.io/gh/TGuillerme/dispRity/branch/master/graph/badge.svg)](https://codecov.io/gh/TGuillerme/dispRity)
[![Project Status: Active - The project has reached a stable, usable state and is being actively developed.](http://www.repostatus.org/badges/latest/active.svg)](http://www.repostatus.org/#active)
<!-- badges out end   -->

## Help

If you need help with the package, hopefully the following manual will be useful.
However, parts of this package are still in development and some other parts are probably not covered.
Thus if you have suggestions or comments on on what has already been developed or will be developed, please send me an email (<guillert@tcd.ie>) or if you are a GitHub user, directly create an issue on the [GitHub page](https://github.com/TGuillerme/dispRity).

## Citations

You can cite both the package or this manual with the following citation:

> Guillerme, T. (2016). dispRity: a package for measuring disparity in R. Zenodo. 10.5281/zenodo.55646

<!-- > Guillerme, T. & Cooper, N. (2016). The dispRity R package manual. Fighsare XXXX -->

Note that this citation is only temporary (but can still be used!).
A proper description of the package is currently in review in Methods in Ecology and Evolution and should be re-submitted this winter.
Upon acceptance (hopefully!), a proper DOI will be released for this manual and the package description.

<!--chapter:end:index.Rmd-->

---
title: "Glossary"
author: "Thomas Guillerme (guillert@tcd.ie) and Natalie Cooper (natalie.cooper@nhm.ac.uk)"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
---

# Glossary

-   **Multidimensional space**. The mathematical multidimensional object that will be analysed with this package.
    In morphometrics, this is often referred to as the morphospace.
    However it may also be referred to as the cladisto-space for cladistic data or the eco-space for ecological data etc.
    In practice, this term designates a matrix where the columns represent the dimensions of the space (often – but not necessarily - > 3!) and the rows represent the elements within this space.

-   **Elements**. The rows of the multidimensional space matrix. Elements can be taxa, field sites, countries etc.

-   **Dimensions**. The columns of the multidimensional space matrix. The dimensions can be referred to as axes of variation, or principal components, for ordinated spaces obtained from a PCA for example.

-   **Subsets**. Subsets of the multidimensional space.
    A subset (or subsets) contains the same number of dimensions as the space but may contain a smaller subset of elements.
    For example, if our space is composed of birds and mammals (the elements) and 50 principal components of variation (the dimensions), we can create two subsets containing just mammals or birds, but with the same 50 dimensions, to compare disparity in the two clades.

<!--chapter:end:01_glossary.Rmd-->

---
title: "Getting started with dispRity"
author: "Thomas Guillerme (guillert@tcd.ie) and Natalie Cooper (natalie.cooper@nhm.ac.uk)"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
---

# Getting started with `dispRity`

## What sort of data does `dispRity` work with?

Disparity can be estimated from pretty much any matrix.
Classically, however, it is measured from ordinated matrices.
These matrices can be from any type of ordination (PCO, PCA, PCoA, MDS, etc.) as long as they have your element names as rows (taxa, experiments, countries etc.) and your ordination axes as columns (the dimensions of your dataset).

### Ordination matrices from `Claddis`

```{r, eval = TRUE, echo = FALSE, message = FALSE}
## Checking if Claddis is installed BUILD ONLY
if(!require(Claddis)) install.packages("Claddis")
```

`dispRity` package can easily take data from `Claddis` using the `Claddis.ordination` function.
For this, simply input a matrix in the `Claddis` format to the function and it will automatically calculate and ordinate the distances among taxa:

```{r}
require(Claddis)

## Ordinating the example data from Claddis
Claddis.ordination(Michaux1989) 
```

Note that several options are available, namely which type of distance should be computed.
See more info in the function manual (`?Claddis.ordination`).
Alternatively, it is of course also possible to manual calculate the ordination matrix using the functions `Claddis::MorphDistMatrix` and `stats::cmdscale`.

### Ordination matrices from `geomorph`

```{r, eval = TRUE, echo = FALSE, message = FALSE}
## Checking if geomorph is installed BUILD ONLY
if(!require(geomorph)) install.packages("geomorph")
```

You can also easily use data from `geomorph` using the `geomorph.ordination` function.
This function simply takes Procrustes aligned data and performs an ordination:

```{r}
require(geomorph)

## Loading the plethodon dataset
data(plethodon)

## Performing a Procrustes transform on the landmarks
procrustes <- gpagen(plethodon$land, PrinAxes = FALSE, print.progress = FALSE)

## Ordinating this data
geomorph.ordination(procrustes)[1:5,1:5]
```

Options for the ordination (from `?prcomp`) can be directly passed to this function to perform customised ordinations.
Additionally you can give the function a `geomorph.data.frame` object.
If the latter contains sorting information (i.e. factors), they can be directly used to make a customised `dispRity` object [customised `dispRity` object](#customised-subsets)!

```{r}
## Using a geomorph.data.frame
geomorph_df <- geomorph.data.frame(procrustes,
     species = plethodon$species, site = plethodon$site)

## Ordinating this data and making a dispRity object
geomorph.ordination(geomorph_df)
```

More about these `dispRity` objects below!

### Other kinds of ordination matrices

If you are not using the packages mentioned above (`Claddis` and `geomorph`) you can easily make your own ordination matrices by using the following functions from the `stats` package.
Here is how to do it for the following types of matrices:

 * Multivariate matrices (principal components analysis; PCA)

```{r}
## A multivariate matrix
head(USArrests)

## Ordinating the matrix using `prcomp` 
ordination <- prcomp(USArrests)

## Selecting the ordinated matrix
ordinated_matrix <- ordination$x
head(ordinated_matrix)
```
  This results in a ordinated matrix with US states as elements and four dimensions (PC 1 to 4). For an alternative method, see the `?princomp` function.

 * Distance matrices (classical multidimensional scaling; MDS)

```{r}
## A matrix of distances between cities
str(eurodist)

## Ordinating the matrix using cmdscale() with k = 5 dimensions 
ordinated_matrix <- cmdscale(eurodist, k = 5)
head(ordinated_matrix)
```

  This results in a ordinated matrix with European cities as elements and five dimensions.

Of course any other method for creating the ordination matrix is totally valid, you can also not use any ordination at all!
The only requirements for the `dispRity` functions is that the input is a matrix with elements as rows and dimensions as columns.

## Performing a simple dispRity analysis

Two `dispRity` functions allow users to run an analysis pipeline simply by inputting an ordination matrix.
These functions allow users to either calculate the disparity through time (`dispRity.through.time`) or the disparity of user-defined groups (`dispRity.per.group`).

**IMPORTANT**

Note that `disparity.through.time` and `disparity.per.group` are wrapper functions (i.e. they incorporate lots of other functions) that allow users to run a basic disparity-through-time, or disparity among groups, analysis without too much effort.
As such they use a lot of default options. 
These are described in the help files for the functions that are used to make the wrapper functions, and not described in the help files for `disparity.through.time` and `disparity.per.group`. 
These defaults are good enough for **data exploration**, but for a proper analysis you should consider the **best parameters for your question and data**. 
For example, which metric should you use?
How many bootstraps do you require?
What model of evolution is most appropriate if you are time slicing?
Should you rarefy the data?
See [`time.subsets`](#time-slicing), [`custom.subsets`](#customised-subsets), [`boot.matrix`](#bootstraps-and-rarefactions) and [`dispRity.metric`](#disparity-metrics) for more details of the defaults used in each of these functions.
Note that any of these default arguments can be changed within the `disparity.through.time` or `disparity.per.group` functions.

### Example data

To illustrate these functions, we will use data from @beckancient2014.
This dataset contains an ordinated matrix of 50 discrete characters from mammals (`BeckLee_mat50`), another matrix of the same 50 mammals and the estimated discrete data characters of their descendants (thus 50 + 49 rows, `BeckLee_mat99`), a dataframe containing the ages of each taxon in the dataset (`BeckLee_ages`) and finally a phylogenetic tree with the relationships among the 50 mammals (`BeckLee_tree`).

```{r, fig.width=7, fig.height=7}
## Loading the ordinated matrices
data(BeckLee_mat50)
data(BeckLee_mat99)

## The first five taxa and dimensions of the 50 taxa matrix
head(BeckLee_mat50[, 1:5])

## The first five taxa and dimensions of the 99 taxa + ancestors matrix
BeckLee_mat99[c(1, 2, 98, 99), 1:5]

## Loading a list of first and last occurrence dates for the fossils
data(BeckLee_ages)
head(BeckLee_ages)

## Loading and plotting the phylogeny
data(BeckLee_tree)
plot(BeckLee_tree, cex = 0.8) 
axisPhylo(root = 140)
nodelabels(cex = 0.5)
```

Of course you can use your own data as detailed in the [previous chapter](#What-sort-of-data-does-dispRity-work-with).

### Disparity through time

The `dispRity.through.time` function calculates disparity through time, a common analysis in palaeontology.
This function (and the following one) uses an analysis pipeline with a lot of default parameters to make the analysis as simple as possible. 
Of course all the defaults can be changed if required, more on this later.

For a disparity through time analysis, you will need:

  * An ordinated matrix (we covered that above)
  * A phylogenetic tree: this must be a `phylo` object (from the `ape` package) and needs a `root.time` element. To give your tree a root time (i.e. an age for the root), you can simply do\\ `my_tree$root.time <- my_age`.
  * The required number of time subsets (here `time = 3`)
  * Your favourite disparity metric (here the sum of variances)

Using the Beck and Lee (2014) data described [above](#example-data):

```{r}
## Measuring disparity through time
disparity_data <- dispRity.through.time(BeckLee_mat50, BeckLee_tree,
                                        time = 3, metric = c(sum, variances))
```

This generates a `dispRity` object (see [here](#guts) for technical details).
When displayed, these `dispRity` objects provide us with information on the operations done to the matrix:

```{r}
## Print the disparity_data object
disparity_data
```

We asked for three subsets (evenly spread across the age of the tree), the data was bootstrapped 100 times (default) and the metric used was the sum of variances.

We can now summarise or plot the `disparity_data` object, or perform statistical tests on it (e.g. a simple `lm`):

```{r, fig.width=7, fig.height=7}
## Summarising disparity through time
summary(disparity_data)

## Plotting the results
plot(disparity_data, type = "continuous")

## Testing for an difference among the time bins
disp_lm <- test.dispRity(disparity_data, test = lm, comparisons = "all")
summary(disp_lm)
```

Please refer to the [specific tutorials](#specific-tutorial) for (much!) more information on the nuts and bolts of the package.
You can also directly explore the specific function help files within R and navigate to related functions.

### Disparity among groups

The `dispRity.per.group` function is used if you are interested in looking at disparity among groups rather than through time.
For example, you could ask if there is a difference in disparity between two groups?

To perform such an analysis, you will need:

 * An matrix with rows as elements and columns as dimensions (always!)
 * A list of group members: this list should be a list of numeric vectors or names corresponding to the row names in the matrix. For example `list("a" = c(1,2), "b" = c(3,4))` will create a group _a_ containing elements 1 and 2 from the matrix and a group _b_ containing elements 3 and 4. Note that elements can be present in multiple groups at once.
 * Your favourite disparity metric (here the sum of variances)

Using the @beckancient2014 data described [above](#example-data):

```{r}
## Creating the two groups (crown versus stem) as a list
mammal_groups <- crown.stem(BeckLee_tree, inc.nodes = TRUE)

## Measuring disparity for each group
disparity_data <- dispRity.per.group(BeckLee_mat50, group = mammal_groups,
                                     metric = c(sum, variances))
```

We can display the disparity of both groups by simply looking at the output variable (`disparity_data`) and then summarising the `disparity_data` object and plotting it, and/or by performing a statistical test to compare disparity across the groups (here a Wilcoxon test).

```{r, fig.width=7, fig.height=7}
## Print the disparity_data object
disparity_data

## Summarising disparity in the different groups
summary(disparity_data)

## Plotting the results
plot(disparity_data)

## Testing for a difference between the groups
test.dispRity(disparity_data, test = wilcox.test, details = TRUE)
```

<!--chapter:end:02_getting-started.Rmd-->

@@

<!--chapter:end:03_specific-tutorials.Rmd-->

---
title: "Data simulation features"
author: "Thomas Guillerme (guillert@tcd.ie) and Natalie Cooper (natalie.cooper@nhm.ac.uk)"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
---

# Making stuff up!

The `dispRity` package also offers some advanced data simulation features to allow to test hypothesis, explore ordinate-spaces or metrics properties or simply playing around with data!
All the following functions are based on the same modular architecture of the package and therefore can be used with most of the functions of the package.

## Simulating discrete morphological data

The function `sim.morpho` allows to simulate discrete morphological data matrices (sometimes referred to as "cladistic" matrices).
It allows to evolve multiple discrete characters on a given phylogenetic trees, given different models, rates, and states.
It even allows to include proper inapplicable data to make datasets as messy as in real life!

In brief, the function `sim.morpho` takes a phylogenetic tree, the number of required characters, the evolutionary model, and a function from which to draw the rates.
The package also contains a function for quickly checking the matrix's phylogenetic signal (as defined in systematics not phylogenetic comparative methods) using parsimony.
The methods are described in details below

```{r}
set.seed(3)
## Simulating a starting tree with 15 taxa as a random coalescent tree
my_tree <- rcoal(15)

## Generating a matrix with 100 characters (85% binary and 15% three state) and
## an equal rates model with a gamma rate distribution (0.5, 1) with no 
## invariant characters.
my_matrix <- sim.morpho(tree = my_tree, characters = 100, states = c(0.85,
    0.15), rates = c(rgamma, 0.5, 1), invariant = FALSE)

## The first few lines of the matrix
my_matrix[1:5, 1:10]

## Checking the matrix properties with a quick Maximum Parsimony tree search
check.morpho(my_matrix, my_tree)
```

Note that this example produces a tree with a great consistency index and an identical topology to the random coalescent tree!
Nearly too good to be true...

### A more detailed description

The protocol implemented here to generate discrete morphological matrices is based on the ones developed in [@GuillermeCooper;@OReilly2016;@puttick2017uncertain;OReilly2017].

 * The first `tree` argument will be the tree on which to "evolve" the characters and therefore requires branch length.
 You can generate quick and easy random Yule trees using `ape::rtree(number_of_taxa)` but I would advise to use more realistic trees for more realistic simulations based on more realistic models (really realistic then) using the function `tree.bd` from the [`diversitree`](http://www.zoology.ubc.ca/prog/diversitree/) package [@fitzjohndiversitree2012].
 * The second argument, `character` is the number of characters. Pretty straight forward.
 * The third, `states` is the proportion of characters states above two (yes, the minimum number of states is two). This argument intakes the proportion of *n*-states characters, for example `states = c(0.5,0.3,0.2)` will generate 50% of binary-state characters, 30% of three-state characters and 20% of four-state characters. There is no limit in the number of state characters proportion as long as the total makes up 100%. 
 * The forth, `model` is the evolutionary model for generating the character(s). More about this below.
 * The fifth and sixth, `rates` and `substitution` are the model parameters described below as well.
 * Finally, the two logical arguments, are self explanatory: `invariant` whether to allow invariant characters (i.e. characters that don't change) and `verbose` whether to print the simulation progress on your console.

#### Available evolutionary models

There are currently three evolutionary models implemented in `sim.morpho` but more will come in the future.
Note also that they allow fine tuning parameters making them pretty plastic!

 * `"ER"`: this model allows any number of character states and is based on the Mk model [@lewisa2001]. It assumes a unique overall evolutionary rate equal substitution rate between character states. This model is based on the `ape::rTraitDisc` function.
 * `"HKY"`: this is binary state character model based on the molecular HKY model [@HKY85]. It uses the four molecular states (A,C,G,T) with a unique overall evolutionary rate and a biased substitution rate towards transitions (A <-> G or C <-> T) against transvertions (A <-> C and G <-> T). After evolving the nucleotide, this model transforms them into binary states by converting the purines (A and G) into state 0 and the pyrimidines (C and T) into state 1. This method is based on the `phyclust::seq.gen.HKY` function and was first proposed by @OReilly2016.
 * `"MIXED"`: this model uses a random (uniform) mix between both the `"ER"` and the `"HKY"` models.

The models can take the following parameters:
(1) `rates` is the evolutionary rate (i.e. the rate of changes along a branch: the evolutionary speed) and (2) `substitution` is the frequency of changes between one state or another.
For example if a character can have high probability of changing (the *evolutionary* rate) with, each time a change occurs a probability of changing from state *X* to state *Y* (the *substitution* rate).
Note that in the `"ER"` model, the substitution rate is ignore because... by definition this (substitution) rate is equal!

The parameters arguments `rates` and `substitution` takes a distributions from which to draw the parameters values for each character.
For example, if you want an `"HKY"` model with an evolutionary rate (i.e. speed) drawn from a uniform distribution bounded between 0.001 and 0.005, you can define it as `rates = c(runif, min = 0.001, max = 0.005)`, `runif` being the function for random draws from a uniform distribution and `max` and `min` being the distribution parameters.
These distributions should always be passed in the format `c(random_distribution_function, distribution_parameters)` with the names of the distribution parameters arguments.

#### Checking the results

An additional function, `check.morpho` runs a quick Maximum Parsimony tree search using the `phangorn` parsimony algorithm.
It quickly calculates the parsimony score, the consistency and retention indices and, if a tree is provided (e.g. the tree used to generate the matrix) it calculates the Robinson-Foulds distance between the most parsimonious tree and the provided tree to determine how different they are.

#### Adding inapplicable characters

Once a matrix is generated, it is possible to apply inapplicable characters to it for increasing realism!
Inapplicable characters are commonly designated as `NA` or simply `-`.
They differ from missing characters `?` in their nature by being inapplicable rather than unknown.
For example, considering a binary character defined as "colour of the tail" with the following states "blue" and "red"; on a taxa with no tail, the character should be coded as inapplicable ("`-`") since the state of the character "colour of tail" is *known*: it's neither "blue" or "red", it's just not there!
It contrasts with coding it as missing ("`?`" - also called as ambiguous) where the state is *unknown*, for example, the taxon of interest is a fossil where the tail has no colour preserved or is not present at all due to bad conservation!

This type of characters can be added to the simulated matrices using the `apply.NA` function/
It takes, as arguments, the `matrix`, the source of inapplicability (`NAs` - more below), the `tree` used to generate the matrix and the two same `invariant` and `verbose` arguments as defined above.
The `NAs` argument allows two types of sources of inapplicability:

 * `"character"` where the inapplicability is due to the character (e.g. coding a character tail for species with no tail).
 In practice, the algorithm chooses a character *X* as the underlying character (e.g. "presence and absence of tail"), arbitrarily chooses one of the states as "absent" (e.g. 0 = absent) and changes in the next character *Y* any state next to character *X* state 0 into an inapplicable token ("`-`").
 This simulates the inapplicability induced by coding the characters (i.e. not always biological).
 * `"clade"` where the inapplicability is due to evolutionary history (e.g. a clade loosing its tail).
 In practice, the algorithm chooses a random clade in the tree and a random character *Z* and replaces the state of the taxa present in the clade by the inapplicable token ("`-`").
 This simulates the inapplicability induced by evolutionary biology (e.g. the lose of a feature in a clade).

To apply these sources of inapplicability, simply repeat the number of inapplicable sources for the desired number of characters with inapplicable data.

```{r}
## Generating 5 "character" NAs and 10 "clade" NAs
my_matrix_NA <- apply.NA(my_matrix, tree = my_tree,
                         NAs = c(rep("character", 5), rep("clade", 10)))

## The first few lines of the resulting matrix
my_matrix_NA[1:10, 90:100]
```
<!--TG: Add inapplicable paper here -->

### Parameters for a realistic(ish) matrix

There are many parameters that can create a "realistic" matrix (i.e. not too different from the input tree with a consistency and retention index close to what is seen in the literature) but because of the randomness of the matrix generation not all parameters combination end up creating "good" matrices.
The following parameters however, seem to generate fairly "realist" matrices with a starting coalescent tree, equal rates model with 0.85 binary characters and 0.15 three state characters, a gamma distribution with a shape parameter ($\alpha$) of 5 and no scaling ($\beta$ = 1) with a rate of 100.

```{r}
set.seed(0)
## tree
my_tree <- rcoal(15)
## matrix
morpho_mat <- sim.morpho(my_tree, characters = 100, model = "ER",
    rates = c(rgamma, rate = 100, shape = 5), invariant = FALSE)
check.morpho(morpho_mat, my_tree)
```

<!-- ## Simulating continuous morphological data -->
<!-- TG: Add tutorial for evolving "morpho" shapes -->

## Simulating multidimensional spaces
Another way to simulate data is to directly simulate an ordinated space with the `space.maker` function.
This function allows users to simulate multidimensional spaces with a certain number of properties.
For example, it is possible to design a multidimensional space with a specific distribution on each axis, a correlation between the axes and a specific cumulative variance per axis.
This can be useful for creating ordinated spaces for null hypothesis, for example if you're using the function `null.test` [@diaz2016global].

This function takes as arguments the number of elements (data points - `elements` argument) and dimensions (`dimensions` argument) to create the space and the distribution functions to be used for each axis.
The distributions are passed through the `distribution` argument as... modular functions!
You can either pass a single distribution function for all the axes (for example `distribution = runif` for all the axis being uniform) or a specific distribution function for each specific axis (for example `distribution = c(runif, rnorm, rgamma))` for the first axis being uniform, the second normal and the third gamma).
You can of course use your very own functions or use the ones implemented in `dispRity` for more complex ones (see below).
Specific optional arguments for each of these distributions can be passed as a list via the `arguments` argument.

Furthermore, it is possible to add a correlation matrix to add a correlation between the axis via the `cor.matrix` argument or even a vector of proportion of variance to be bear by each axis via the `scree` argument to simulate realistic ordinated spaces.

Here is a simple two dimensional example:

```{r, fig.width=8, fig.height=8, eval=TRUE}
## Graphical options
op <- par(bty = "n")

## A square space
square_space <- space.maker(100, 2, runif)

## The resulting 2D matrix
head(square_space)

## Visualising the space
plot(square_space, pch = 20, xlab = "", ylab = "", main = "Uniform 2D space")
```

Of course, more complex spaces can be created by changing the distributions, their arguments or adding a correlation matrix or a cumulative variance vector:

```{r}
## A plane space: uniform with one dimensions equal to 0
plane_space <- space.maker(2500, 3, c(runif, runif, runif),
                           arguments = list(list(min = 0, max = 0), NULL, NULL))

## Correlation matrix for a 3D space
(cor_matrix <- matrix(cbind(1, 0.8, 0.2, 0.8, 1, 0.7, 0.2, 0.7, 1), nrow = 3))

## An ellipsoid space (normal space with correlation)
ellipse_space <- space.maker(2500, 3, rnorm, cor.matrix = cor_matrix)

## A cylindrical space with decreasing axes variance
cylindrical_space <- space.maker(2500, 3, c(rnorm, rnorm, runif),
                                 scree = c(0.7, 0.2, 0.1))
```

### Personalised dimensions distributions

Following the modular architecture of the package, it is of course possible to pass home made distribution functions to the `distribution` argument.
For example, the `random.circle` function is a personalised one implemented in `dispRity`.
This function allows to create circles based on basic trigonometry allowing to axis to covary to produce circle coordinates.
By default, this function generates two sets of coordinates with a `distribution` argument and a minimum and maximum boundary (`inner` and `outer` respectively) to create nice sharp edges to the circle.
The maximum boundary is equivalent to the radius of the circle (it removes coordinates beyond the circle radius) and the minimum is equivalent to the radius of a smaller circle with no data (it removes coordinates below this inner circle radius).

```{r, fig.width=8, fig.height=8, eval=TRUE}
## Graphical options
op <- par(bty = "n")

## Generating coordinates for a normal circle with a upper boundary of 1
circle <- random.circle(1000, rnorm, inner = 0, outer = 1)

## Plotting the circle
plot(circle, xlab = "x", ylab = "y", main = "A normal circle")

## Creating doughnut space (a spherical space with a hole)
doughnut_space <- space.maker(5000, 3, c(rnorm, random.circle),
     arguments = list(list(mean = 0), list(runif, inner = 0.5, outer = 1)))
```

### Visualising the space

I suggest using the excellent `scatterplot3d` package to play around and visualise the simulated spaces:

```{r, fig.width=8, fig.height=8, eval=TRUE}
## Graphical options
op <- par(mfrow = (c(2, 2)), bty = "n")
## Visualising 3D spaces
require(scatterplot3d)

## The plane space
scatterplot3d(plane_space, pch = 20, xlab = "", ylab = "", zlab = "",
              xlim = c(-0.5, 0.5), main = "Plane space")

## The ellipsoid space
scatterplot3d(ellipse_space, pch = 20, xlab = "", ylab = "", zlab = "",
              main = "Normal ellipsoid space")

## A cylindrical space with a decreasing variance per axis
scatterplot3d(cylindrical_space, pch = 20, xlab = "", ylab = "", zlab = "",
              main = "Normal cylindrical space")
## Axes have different orders of magnitude

## Plotting the doughnut space
scatterplot3d(doughnut_space[,c(2,1,3)], pch = 20, xlab = "", ylab = "",
              zlab = "", main = "Doughnut space")
par(op)
```

### Generating realistic spaces

It is possible to generate "realistic" spaces by simply extracting the parameters of an existing space and scaling it up to the simulated space.
For example, we can extract the parameters of the `BeckLee_mat50` ordinated space and simulate a similar space.

```{r, fig.width=8, fig.height=8, eval=TRUE}
## Loading the data
data(BeckLee_mat50)

## Number of dimensions
obs_dim <- ncol(BeckLee_mat50)

## Observed correlation between the dimensions
obs_correlations <- cor(BeckLee_mat50)

## Observed mean and standard deviation per axis
obs_mu_sd_axis <- mapply(function(x,y) list("mean" = x, "sd" = y),
                         as.list(apply(BeckLee_mat50, 2, mean)),
                         as.list(apply(BeckLee_mat50, 2, sd)), SIMPLIFY = FALSE)

## Observed overall mean and standard deviation
obs_mu_sd_glob <- list("mean" = mean(BeckLee_mat50), "sd" = sd(BeckLee_mat50))

## Scaled observed variance per axis (scree plot)
obs_scree <- variances(BeckLee_mat50)/sum(variances(BeckLee_mat50))

## Generating our simulated space
simulated_space <- space.maker(1000, dimensions = obs_dim, 
                               distribution = rep(list(rnorm), obs_dim),
                               arguments = obs_mu_sd_axis,
                               cor.matrix = obs_correlations)

## Visualising the fit of our data in the space (in the two first dimensions)
plot(simulated_space[,1:2], xlab = "PC1", ylab = "PC2")
points(BeckLee_mat50[,1:2], col = "red", pch = 20)
legend("topleft", legend = c("observed", "simulated"),
        pch = c(20,21), col = c("red", "black"))
```

It is now possible to simulate a space using these observed arguments to test several hypothesis:

 * Is the space uniform or normal?
 * If the space is normal, is the mean and variance global or specific for each axis?

```{r, fig.width=8, fig.height=8, eval=TRUE}
## Measuring disparity as the sum of variance
observed_disp <- dispRity(BeckLee_mat50, metric = c(median, centroids))

## Is the space uniform?
test_unif <- null.test(observed_disp, null.distrib = runif)

## Is the space normal with a mean of 0 and a sd of 1?
test_norm1 <- null.test(observed_disp, null.distrib = rnorm)

## Is the space normal with the observed mean and sd and cumulative variance
test_norm2 <- null.test(observed_disp, null.distrib = rep(list(rnorm), obs_dim),
                        null.args = rep(list(obs_mu_sd_glob), obs_dim),
                        null.scree = obs_scree)

## Is the space multiple normal with multiple means and sds and a correlation?
test_norm3 <- null.test(observed_disp, null.distrib = rep(list(rnorm), obs_dim),
                        null.args = obs_mu_sd_axis, null.cor = obs_correlations)

## Graphical options
op <- par(mfrow = (c(2, 2)), bty = "n")
## Plotting the results
plot(test_unif, main = "Uniform (0,1)")
plot(test_norm1, main = "Normal (0,1)")
plot(test_norm2, main = paste0("Normal (", round(obs_mu_sd_glob[[1]], digit = 3),
                              ",", round(obs_mu_sd_glob[[2]], digit = 3), ")"))
plot(test_norm3, main = "Normal (variable + correlation)")

If we measure disparity as the median distance from the morphospace centroid, we can explain the distribution of the data as normal with the variable observed mean and standard deviation and with a correlation between the dimensions.

<!--chapter:end:04_making-up-data.Rmd-->

---
title: "The guts of the dispRity package"
author: "Thomas Guillerme (guillert@tcd.ie)"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
---

# The guts of the `dispRity` package

## Manipulating `dispRity` objects

Disparity analysis involves a lot of manipulation of many matrices (especially when bootstrapping) which can be impractical to visualise and will quickly overwhelm your `R` console.
Even the simple Beck and Lee 2014 example above produces an object with > 72 lines of lists of lists of matrices!

Therefore `dispRity` uses a specific class of object called a `dispRity` object.
These objects allow users to use S3 method functions such as `summary.dispRity`, `plot.dispRity` and `print.dispRity`. 
`dispRity` also contains various utility functions that manipulate the `dispRity` object (e.g. `sort.dispRity`, `extract.dispRity` see the full list in the next section).
These functions modify the `dispRity` object without having to delve into its complex structure!
The full structure of a `dispRity` object is detailed [here](https://github.com/TGuillerme/dispRity/blob/master/disparity_object.md).

```{r}
## Loading the example data
data(disparity)

## What is the class of the median_centroids object?
class(disparity)

## What does the object contain?
names(disparity)

## Summarising it using the S3 method print.dispRity
disparity
```

Note that it is always possible to recall the full object using the argument `all = TRUE` in `print.dispRity`:

```{r, eval=FALSE}
## Display the full object
print(disparity, all = TRUE)
## This is more nearly ~ 5000 lines on my 13 inch laptop screen!
```

## `dispRity` utilities

The package also provides some utility functions to facilitate multidimensional analysis.

### `dispRity` object utilities  <a name="dispRity.utilities"></a>
The first set of utilities are functions for manipulating `dispRity` objects:

#### `make.dispRity`
This function creates empty `dispRity` objects.

```{r}
## Creating an empty dispRity object
make.dispRity()

## Creating an "empty" dispRity object with a matrix
(disparity_obj <- make.dispRity(matrix(rnorm(20), 5, 4)))
```

####  `fill.dispRity`
This function initialises a `dispRity` object and generates its call properties.

```{r}
## The dispRity object's call is indeed empty
disparity_obj$call

## Filling an empty disparity object (that needs to contain at least a matrix)
(disparity_obj <- fill.dispRity(disparity_obj))

## The dipRity object has now the correct minimal attributes
disparity_obj$call
```

####  `matrix.dispRity`
This function extracts a specific matrix from a disparity object.
The matrix can be one of the bootstrapped matrices or/and a rarefied matrix.

```{r}
## Extracting the matrix containing the coordinates of the elements at time 50
str(matrix.dispRity(disparity, "50"))

## Extracting the 3rd bootstrapped matrix with the 2nd rarefaction level
## (15 elements) from the second group (80 Mya)
str(matrix.dispRity(disparity, subsets = 1, bootstrap = 3, rarefaction = 2))
```

#### `get.subsets.dispRity`
This function creates a dispRity object that contains only elements from one specific subsets.

```{r, eval=FALSE}
## Extracting all the data for the crown mammals
(crown_mammals <- get.subsets.dispRity(disp_crown_stemBS, "Group.crown"))

## The object keeps the properties of the parent object but is composed of only one subsets
length(crown_mammals$subsets)
```

#### `extract.dispRity`
This function extracts the calculated disparity values of a specific matrix.

```{r, eval=FALSE}
## Extracting the observed disparity (default)
extract.dispRity(disparity)

## Extracting the disparity from the bootstrapped values from the
## 10th rarefaction level from the second subsets (80 Mya)
extract.dispRity(disparity, observed = FALSE, subsets = 2, rarefaction = 10)
```

#### `scale.dispRity`
This is the S3 method of `scale` (scaling and/or centring) that can be applied to the disparity data of a `dispRity` object.

```{r, eval=FALSE}
## Getting the disparity values of the time subsets
head(summary(disparity))

## Scaling the same disparity values
head(summary(scale(disparity, scale = TRUE)))

## Scaling and centering:
head(summary(scale(disparity, scale = TRUE, center = TRUE)))
```

#### `sort.dispRity`
This is the S3 method of `sort` for sorting the subsets alphabetically (default) or following a specific pattern.

```{r, eval=FALSE}
## Sorting the disparity subsets in inverse alphabetic order
head(summary(sort(disparity, decreasing = TRUE)))

## Customised sorting
head(summary(sort(disparity, sort = c(7, 1, 3, 4, 5, 2, 6))))
```

## The `dispRity` object content
The functions above are utilities to easily and safely access different elements in the `dispRity` object.
Alternatively, of course, each elements can be accessed manually.
Here is an explanation on how it works.
The `dispRity` object is a `list` of two to four elements, each of which are detailed below:
